### 2013-07-19 doc/asplos14/conclude.tex
make asplos14 directory from sosp13

Irregular computations are both important and challenging to execute quickly.
Scaling these applications easily to commodity hardware has been a historical
challenge. \Grappa is a runtime framework that simplifies this task for
software developers and compiler writers. This paper describes the \Grappa
framework and its three main components: a task library, a distributed shared
memory system, and a network aggregator to make commodity networks efficient
with small message sizes. \Grappa's key aspect is extreme latency tolerance,
which not only hides network latency but also enables the system to spend time
on sophisticated work stealing and network optimizations, trading latency for
even more throughput.

Our evaluation of \Grappa reveals that the core components, scheduling and communication, achieve their design goals.  Thousands of workers can be efficiently context switched on a multicore processor, up to the DRAM bandwidth.  Aggregating messages enables \Grappa to achieve over 1.0 GUPS on 64 nodes.  We also explored four other algorithms: unbalanced tree search, breadth first search, PageRank, and integer sort, comparing performance of these algorithms on \Grappa, the Cray XMT, and against hand-optimized MPI implementations.  As would be expected, when the MPI implementation exploits in the application, many of the same techniques \Grappa provides in the runtime, such as aggregating requests, yet without the overhead of tasking, the MPI implementation excels.  This comes with enormous implementation complexity burdened on the application developer, however, and significant tuning has yet to be done on the \Grappa runtime system.  Compared to the Cray XMT \Grappa is between 2.6X faster and 4.3X slower.  Yet when cost-performance is considered, mass-market x86 cluster hardware makes \Grappa a highly attractive option.

>>>>

Irregular computations are both important and challenging to execute quickly.
Scaling these applications easily on commodity hardware has been a historical
challenge. \Grappa is a runtime framework that simplifies this task for
software developers and compiler writers. This paper describes the \Grappa
framework and its three main components: a task library, a distributed shared
memory system, and a network aggregator. \Grappa's key aspect is extreme latency tolerance,
which not only hides network latency but also enables the system to spend time
on sophisticated work stealing and network optimizations, trading latency for
even more throughput.

Our evaluation of \Grappa reveals that the core components, scheduling and communication, achieve their design goals.  Thousands of workers can be efficiently context switched on a multicore processor, up to the DRAM bandwidth.  Aggregating messages enables \Grappa to achieve over 1.0 GUPS on 64 nodes.  We also explored four other algorithms: unbalanced tree search, breadth first search, PageRank, and integer sort, comparing performance of these algorithms on \Grappa, the Cray XMT, and hand-optimized MPI implementations.  As would be expected, when the MPI implementation exploits in the application, many of the same techniques \Grappa provides in the runtime, such as aggregating requests, yet without the overhead of tasking, the MPI implementation excels.  This comes with enormous implementation complexity, burdening the application developer. Moreover, significant tuning has yet to be done on the \Grappa runtime system.  Compared to the Cray XMT \Grappa is between 2.6X faster and 4.3X slower.  Yet when cost-performance is considered, mass-market x86 cluster hardware makes \Grappa a highly attractive option.

### 2013-07-19 doc/asplos14/evaluation.tex
make asplos14 directory from sosp13

\TODO{Grappa is well-suited to applications where there is
    abundant data parallelism but unpredictable or imbalanced data
    access. BSP is efficient where there is sufficient parallel slack.
    Although Grappa loses some asynchrony by relying heavily on
    aggregation, it still provides more flexibility for sections of the
    computation to proceed more quickly.

@@

As discussed earlier, fast context switching is at the heart of \Grappa's
latency tolerance abilities. We assess context switch overheads using a simple
microbenchmark that has a configurable number of workers, where each worker 
just increments values in a large array.

@@

Figure~\ref{fig:context-switch-exp} shows the average context switch time as
the number of workers grow. There are two important takeaway points from these
data. Context switch time for the number of workers we use ($\approx$1K workers) is
on the order of 50 ns; and prefetching has significant impact on context
switch time. And context switch time does not grow significantly with the
number of workers --- even with one-half million workers, context
switch time is around 75ns. In comparison, for the same yield test using kernel-level
Pthreads on a single core, the switch time is 450ns for a few threads and 800ns for
1000-32000 threads.

If we calculate aggregate context switch\emph{rate} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we emprically found that 4 cachelines (1 for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8 cacheline transfers. The off-chip
bandwidth of a single socket in our system is 270M cachelines per
second \cite{porterfield:bw,Nelson:hotpar11-real}. This implies that
we can sustain at most 34M context switches per second per socket once
the space of all contexts is enough to mostly reside outside of L3.

@@

\paragraph{Global memory and communication.} We measure the performance of
\Grappa's global memory and communication layers using a faithful
implementation of the giga updates per second (GUPs) benchmark.
Read-modify-write updates are dispatched at random to a large array. This
benchmark stresses the communication layer of \Grappa separately from the
scheduler, because only a single worker is used per system node.
Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain well over a
billion updates per second with 64 nodes. Note also that when aggregation is turned off, the update rate is flat at the bottom, demonstrating that aggregation is instrumental for good communication performance.

This compares very favorably to published results~\cite{gups} for other
high-end HPC systems. Though the actual computation done by GUPS is completely
non-sensical, irregular, data-intensive applications typically must be able to
sustain a high rate of random accesses in order to, for example, visit and
mark vertices during a graph traversal. High random access rate in a
distributed setting has been a long-standing challenge in HPC.

@@

\paragraph{Putting it all together with Unbalanced Tree Search in memory
(UTS).} 
Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa running
UTS. The point of this experiment is to demonstrate whether \Grappa's context
switching and communication layers can in fact be used together, while
balancing workload, to run an irregular application efficiently. Visiting
vertices in the distributed tree requires mostly remote accesses, and because
each vertex in the tree must be visited before it can be expanded, blocking
remote reads are required. 
In this case, we are forced to context switch to tolerate the remote access and continue aggregating. 
We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput --- at the 16-node data point,
the average active tasks per core over the search is 775 and 13 for T1 and T3,
respectively. On the other hand \Grappa performs and scales very well for T1
trees.

@@

Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
within striking distance of XMT and well beyond UPC performance (UTS only).
However, it underperforms compared to hand-tuned MPI. Nevertheless, this
comparison needs to be taken with a grain of salt for several reasons. First,
we have not spent as much time tuning \Grappa's implementations. This is
supported by inspecting the BFS MPI implementation we used, which clearly
shows that it employs a lot of algorithmic-specific optimizations. In fact,
some of these optimizations resemble some of what \Grappa does automatically,
like message aggregation. Second, the GUPS results presented earlier suggests
that \Grappa performance can do a lot better.

>>>>

\TODO{Grappa is well-suited to applications where there is
    abundant data parallelism but unpredictable or imbalanced data
    access. BSP is efficient where there is sufficient parallel slack.
    Although Grappa loses some asynchrony by relying heavily on
    aggregation, it still provides more flexibility for sections of the
    computation to proceed more quickly. }

@@

Fast context switching is at the heart of \Grappa's
latency tolerance abilities. We assess context switch overheads using a simple
microbenchmark that runs a configurable number of workers on a single core, where each worker 
increments values in a large array.

@@

Figure~\ref{fig:context-switch-exp} shows the average context switch
time as the number of workers grow. At our standard operating point
($\approx$1K workers), context switch time is on the order of 50
ns. As we add workers, the time increases slowly, but levels off: we
also ran with 500,000 workers (10 times what is shown in the figure)
and found that context switch time was around 75 ns. In comparison,
for the same yield test using kernel-level Pthreads on a single core,
the switch time is 450ns for a few threads and 800ns for 1000-32000
threads.

If we calculate aggregate context switch \emph{rate} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we empirically found that 4 cache lines (1 for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8 cache line transfers. The off-chip
bandwidth of a single socket in our system is 270M cache lines per
second \cite{porterfield:bw,Nelson:hotpar11}. This implies that
we can sustain at most 34M context switches per second per socket once
the space of all contexts is enough to mostly reside outside of L3.

@@

\paragraph{Global memory and communication.} We measure the performance of
\Grappa's global memory and communication layers using a faithful
implementation of the giga updates per second (GUPs) benchmark, which measures cluster-wide random access bandwidth.
Read-modify-write updates are dispatched at random to a large array. This
benchmark stresses the communication layer of \Grappa separately from the
scheduler, because only a single worker is used per system node.
Figure~\ref{fig:grappa-gups} shows that \Grappa is able to sustain well over a
billion updates per second with 64 nodes. Note also that when aggregation is turned off, the update rate is nearly flat. Aggregation is instrumental for good communication performance.

This compares very favorably to published results~\cite{gups} for other
high-end HPC systems. Though the actual computation done by GUPS is not 
useful, irregular, data-intensive applications typically must be able to
sustain a high rate of random accesses in order to, for example, visit and
mark vertices during a graph traversal. High random access rate in a
distributed setting has been a long-standing challenge in HPC.

@@

\paragraph{Putting it all together with Unbalanced Tree Search in memory
(UTS).} 
Figure~\ref{fig:grappa-uts} shows the overall performance of \Grappa running
UTS. This experiment demonstrates that \Grappa's context
switching and communication layers can be used together, while
balancing workload, to run an irregular application efficiently.

Visiting vertices in the distributed tree requires mostly remote
accesses, and because each vertex in the tree must be visited before
it can be expanded, blocking remote reads are required. In this case,
we are forced to context switch to tolerate the remote access and
continue aggregating.

We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput --- at the 16-node data point,
the average active tasks per core over the search is 775 and 13 for T1 and T3,
respectively. On the other hand, \Grappa performs and scales very well for T1
trees.

@@

Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
within striking distance of XMT and well beyond UPC performance (UTS only).
However, it underperforms compared to hand-tuned MPI. We believe there are two reasons for this:
First, Grappa is young; much optimization and tuning of the runtime
and the application is yet to be done. Second, the MPI implementations
we used employ algorithmic-specific optimizations---some of which
resemble of what \Grappa does automatically, like message aggregation.
Grappa's GUPS results presented earlier suggest there is significant
room for improvement.

### 2013-07-19 doc/asplos14/intro.tex
make asplos14 directory from sosp13

The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of
supporting irregular applications by using concurrency to hide latencies. It
had a large distributed shared memory with no caches. On every clock cycle,
each processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully tolerate memory access
latency. The network was designed with a single-word injection rate that
matched the processor clock frequency and sufficient bandwidth to sustain a
reference from every processor on every clock cycle. Unfortunately, the MTA
was not general enough nor cost-effective. The Cray XMT approximates the Tera
MTA-2, reducing its cost but not overcoming its narrow range of applicability.

@@

Our evaluation of \Grappa shows that its core components provide scalable performance.  It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.  The communication layer helps \Grappa achieve over 1.0 GUPS with 64 system nodes.  When we compare \Grappa to native MPI implementations of more complex application kernels, the story is more muddled.  When the MPI implementation is carefully tuned to aggregate messages inside the application -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between 6-10X faster.  We see this in our comparisons for breadth first search and integer sorting.  On the other hand, if the MPI version is implemented in a straightforward manner, as we see with PageRank, GUPS, and unbalanced tree search, \Grappa is 2-5X faster.  When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.  Overall we find this encouraging because \Grappa is scalable and achieves this performance from a mass-market cluster, hence adding more system nodes is an inexpensive option for tackling larger problem sizes.

>>>>

The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of
supporting irregular applications by using concurrency to hide latencies. It
had a large distributed shared memory with no caches. On every clock cycle,
each processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully tolerate memory access
latency. The network was designed with a single-word injection rate that
matched the processor clock frequency and sufficient bandwidth to sustain a
reference from every processor on every clock cycle. Unfortunately, the MTA's relatively low single-threaded performance meant that it
was not general enough nor cost-effective. The Cray XMT approximates the Tera
MTA-2, reducing its cost but not overcoming its narrow range of applicability.

@@

Our evaluation of \Grappa shows that its core components provide scalable performance for irregular applications.  It can multiplex thousands of workers on a multicore CPU and is limited only by DRAM bandwidth, not latency, to fetch worker state.  The communication layer helps \Grappa achieve over 1.0 GUPS with 64 system nodes.  When we compare \Grappa to native MPI implementations of more complex application kernels, the story is more muddled.  When the MPI implementation is carefully tuned to aggregate messages inside the application -- essentially duplicating the work \Grappa is doing for the programmer, without the generality and context switch overhead -- then the MPI implementation is between 6-10X faster.  We see this in our comparisons for breadth first search and integer sorting.  On the other hand, if the MPI version is implemented in a straightforward manner, as we see with PageRank, GUPS, and unbalanced tree search, \Grappa is 2-5X faster.  When we compare \Grappa to the custom Cray XMT, we find \Grappa to be between 2.6X faster and 4.3X slower depending on the benchmark.

### 2013-07-19 doc/asplos14/tasks.tex
make asplos14 directory from sosp13

Since \Grappa keeps a very large number of active workers, their context data
will not fit in cache. By oversubscribing on the number of workers
beyond what is required for latency tolerance, the scheduler can
ensure there is always some number of context pointers in the ready
queue. This allows the scheduler to prefetch contexts into
cache. The reason this works is because the size of the L1 cache is
sufficient to hold enough contexts to tolerate the latency to main
memory. Empirically we find that prefetching the fourth worker in the scheduling order is
sufficient. This prefetching makes context switching effectively free of cache misses even to hundreds
of thousands of threads. We provide an analysis of our context switch
performance in Section~\ref{eval:basic}.

@@

\Grappa programmers focus on expressing as much parallelism as possible
without concern for where it will execute. \Grappa then chooses where and when
to exploit this parallelism, scheduling as much work as is necessary on each
core to keep it busy in the presence of system latencies and task dependences.

>>>>

Since \Grappa keeps a very large number of active workers, their context data
will not fit in cache. By oversubscribing on the number of workers
beyond what is required for latency tolerance, the scheduler can
ensure there is always some number of context pointers in the ready
queue. This allows the scheduler to prefetch contexts into
cache using software prefetch instructions. The reason this works is because the size of the L1 cache is
sufficient to hold enough contexts to tolerate the latency to main
memory. Empirically we find that prefetching the fourth worker in the scheduling order is
sufficient. This prefetching makes context switching effectively free of cache misses even to hundreds
of thousands of threads. We provide an analysis of our context switch
performance in Section~\ref{eval:basic}.

@@

\TODO{addressing reviewer2: make it clear that yes Grappa is a low
    level programming model, but it provides the mechanisms for
    implementing a variety of execution models. Further, in some cases
    like the parallel loops, we have allowed for may parallelism to
    give flexibility to the implementation.}
\Grappa programmers focus on expressing as much parallelism as possible
without concern for where it will execute. \Grappa then chooses where and when
to exploit this parallelism, scheduling as much work as is necessary on each
core to keep it busy in the presence of system latencies and task dependences.
